---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "# "
)
```

[![Build Status](https://travis-ci.org/cstubben/tidypmc.svg?branch=master)](https://travis-ci.org/cstubben/tidypmc)
[![Coverage status](https://codecov.io/gh/cstubben/tidypmc/branch/master/graph/badge.svg)](https://codecov.io/github/cstubben/tidypmc?branch=master)
[![](https://badges.ropensci.org/290_status.svg)](https://github.com/ropensci/software-review/issues/290)

# tidypmc

The [Open Access subset] of [Pubmed Central] (PMC) includes 2.5 million articles
from biomedical and life sciences journals.  The full text XML files are freely
available from the [REST service] or [FTP site] for text mining but can be
challenging to parse. For example, section tags are nested to arbitrary depths,
formulas and tables may return incomprehensible text blobs and superscripted
references are pasted at the end of words.  The functions in the `tidypmc`
package are intended to return readable text and maintain the document
structure, so gene names and other terms can be associated with specific
sections, paragraphs, sentences or table rows.


## Installation

Use `devtools` to install the package.

```{r install, eval=FALSE}
devtools::install_github("cstubben/tidypmc")
```

## Load XML

Download a single XML document like [PMC2231364] from the [REST service] using
the `pmc_xml` function.


```{r pmc_xml, echo=-1}
options(width=100)
doc <- pmc_xml("PMC2231364")
doc
```


The [europepmc] package includes additional functions to search PMC
and download full text.  Be sure to include the `OPEN_ACCESS` field in
the search since these are the only articles with full text XML available.

```{r epmc, echo=-1}
options(width=100)
library(europepmc)
yp <- epmc_search("title:(Yersinia pestis virulence) OPEN_ACCESS:Y")
yp[, c(4,11,6)] %>%
  print(n=5)
```


Save the results to a list of XML documents.

```{r purrr, eval=FALSE}
docs <- purrr::map(yp$pmcid, epmc_ftxt)
```


See the [PMC FTP vignette] for details on parsing the large XML files on the [FTP site]
with 10,000 articles each.


## Parse XML


The package includes five functions to parse the `xml_document`.


|R function     |Description                                                                |
|:--------------|:--------------------------------------------------------------------------|
|`pmc_text`     |Split section paragraphs into sentences with full path to subsection titles|
|`pmc_caption`  |Split figure, table and supplementary material captions into sentences     |
|`pmc_table`    |Convert table nodes into a list of tibbles                                 |
|`pmc_reference`|Format references cited into a tibble                                      |
|`pmc_metadata` |List journal and article metadata in front node                            |


The `pmc_text` function uses the [tokenizers] package to split section paragraphs into
sentences.  The function also removes any tables, figures or formulas that are nested
within paragraph tags, replaces superscripted references with brackets, adds carets and
underscores to other superscripts and subscripts and includes the full path to the
subsection title.

```{r pmc_text, echo=-1}
options(width=110)
library(tidypmc)
txt <- pmc_text(doc)
txt
dplyr::count(txt, section, sort=TRUE)
```


Load the [tidytext] package for further text processing.

```{r tidytext, echo=-1}
options(width=110)
library(tidytext)
library(tidyverse)
x1 <- unnest_tokens(txt, word, text) %>%
  anti_join(stop_words) %>%
  filter(!word %in% 1:100)
filter(x1, str_detect(section, "^Results"))
filter(x1, str_detect(section, "^Results")) %>%
  dplyr::count(word, sort = TRUE)
```



The `pmc_table` function formats tables by collapsing multiline headers,
expanding rowspan and colspan attributes and adding subheadings into a new column.

```{r pmc_table, echo=-1}
options(width=110)
tbls <- pmc_table(doc)
map_int(tbls, nrow)
tbls[[1]]
```

Use `collapse_rows` to join column names and cell values in a semi-colon delimited string (and
then search using functions in the next section).

```{r collapserows, echo=-1}
options(width=110)
collapse_rows(tbls, na.string="-")
```

The other three `pmc` functions are described in the package [vignette].


## Searching text

There are a few functions to search within the `pmc_text` or collapsed
`pmc_table` output.  `separate_text` uses the [stringr] package to extract any
regular expression or vector of words.


```{r separate_text, echo=-1}
options(width=110)
separate_text(txt, "[ATCGN]{5,}")
```

A few wrappers search pre-defined patterns and add an extra step to expand
matched ranges. `separate_refs` matches references within brackets using
`\\[[0-9, -]+\\]` and expands ranges like `[7-9]`.

```{r separate_refs, echo=-1}
options(width=110)
separate_refs(txt)
```

`separate_genes` will find microbial genes like tauD (with a
capitalized 4th letter)  and expand operons like `tauABCD` into
four genes.  `separate_tags` will find and expand locus tag ranges below.


```{r locus_tags, echo=-1}
options(width=110)
collapse_rows(tbls, na="-") %>%
  separate_tags("YPO") %>%
  filter(id =="YPO1855")
```


See the [vignette] for more details including code to parse
XML documents using the [xml2] package.  The [PMC FTP vignette]
has details on parsing XML files at the Europe PMC [FTP site].


### Community Guidelines

This project is released with a [Contributor Code of Conduct](CONDUCT.md). By
participating in this project you agree to abide by its terms. Feedback, bug
reports, and feature requests are welcome
[here](https://github.com/cstubben/tidypmc/issues).



[xml2]: https://github.com/r-lib/xml2
[PMC2231364]: https://www.ebi.ac.uk/europepmc/webservices/rest/PMC2231364/fullTextXML
[Open Access subset]: https://europepmc.org/downloads/openaccess
[REST service]: https://europepmc.org/RestfulWebService
[FTP site]: https://europepmc.org/ftp/oa/
[tidytext]: https://www.tidytextmining.com/
[stringr]: https://stringr.tidyverse.org/
[vignette]: https://github.com/cstubben/tidypmc/blob/master/vignettes/tidypmc.md
[PMC FTP vignette]: https://github.com/cstubben/tidypmc/blob/master/vignettes/pmcftp.md
[tokenizers]: https://lincolnmullen.com/software/tokenizers/
[xml2]: https://github.com/r-lib/xml2
[europepmc]: https://github.com/ropensci/europepmc
[Pubmed Central]: https://europepmc.org
